### 2024-03-29 doc/sections/introduction.tex
[ tex ] introduction & contribution

Horne~\cite{Horne15:mav} gave a syntactic proof of cut-elimination for MAV. We present an alternative proof of cut-elimination via a semantic model. This proof avoids some of the intrincate syntactic reasoning in Horne's proof and is more modular, as shown by the extensions we present in \cref{sec:mav-extensions}.

@@

>>>>

\emph{Cut-elimination}, or \emph{admissibility of Cut}, is the fundamental property of Gentzen Sequent Calculi systems, which states that proofs using the Cut rule to introduce ``detours'' can be normalised to ones without. Crucial properties such as consistency and decidability follow directly from Cut-elimination. The Deep Inference analogue of Cut-elimination is the admissibility of the whole \emph{up} fragment of the calculus, which includes the Deep Inference form of Cut (which we describe in \Cref{sec:mav-syntax} below) as well as duals of most of the other rules of the calculus. Admissibility of the up fragment has the same metatheoretic benefits for Deep Inference systems as it does for Sequent Calculus ones.

Guglielmi \cite[\S4.1]{Guglielmi14:di} proves admissibility via the Splitting Theorem, which shows that proofs of conjoined structures can be split into separate subproofs. This is proved by a detailed syntactic analysis of proofs. Horne~\cite{Horne15:mav} gave a syntactic proof of the admissibility of the up fragment for MAV, that extends Guglielmi's technique with further reasoning about the additives. The proof is quite lengthy and involves intrincate syntactic reasoning and the subtle and complex termination measures.

We present an alternative proof of cut-elimination via a semantic model. This proof avoids some of the in Horne's proof. We believe our proof to be more robust in the presence of extensions, due to our use of standard constructions such as Day monoids, order ideals, and the Chu construction.

@@

To our knowledge, all prior work on the metatheory of deep inference systems like BV and MAV has been carried out using syntactic techniques such as rewriting with termination measures, and translations into other logics with known cut-elimination properties. Our main contribution is the use of semantic techniques to derive the admissibility of identity expansion, cut, and the other \emph{co-} rules of MAV. To this end, we develop a number of results on the semantics of BV and MAV:
\begin{enumerate}
\item In \Cref{sec:mav-algebras}, we propose MAV-algebras as the algebraic counterpart of MAV. In short, an MAV-algebra is a $*$-autonomous partial order with meets, and another partially ordered monoidal structure that is \emph{duoidal} with respect to the $*$-autonomous structure.
\item Normal proofs (our name for the ``up'' fragment) as we define them in the next section do not {\it a priori} support all the structure of an MAV-algebra, so we define the weaker notion of \emph{MAV-frame} (\Cref{defn:mav-frame}). These play the same role as Kripke structures do for modal and substructural logics. The first main technical contribution of the paper is in showing that every MAV-frame can be completed to an MAV-algebra.
\item We then show that MAV-frames are strongly complete (in the terminology of Okada) for MAV by proving that the MAV-frame constructed from normal proofs can be used to deduce that all MAV-provable structures have normal proofs in \Cref{thm:cut-elim}. Completeness in the usual sense also follows in \Cref{thm:completeness}.
\end{enumerate}

We describe the MAV Deep Inference system in \Cref{sec:mav-syntax} and motivate the idea for readers not familiar with such systems. We present our proof in the full generality of MAV, but note that essentially the same proof applies to BV as well. We discuss further possible extensions in \Cref{sec:future-work}.

### 2024-03-29 doc/sections/mav-semantics.tex
[ tex ] introduction & contribution

To our knowledge, the semantics of MAV have not been previously
studied in the literature. The rules are an extension of MALL, so an
\emph{MAV-algebra} will partly be a $*$-autonomous partial order
(\Cref{defn:star-autonomous}) with meets and joins. The
additional structure for seq satisfies the
conditions of a duoidal category \cite{hopf-book} (\Cref{defn:duoidal}). We show that MAV is sound for MAV-algebras in
\Cref{thm:soundness}.

@@

BV and MAV extend linear logic by adding a non-commutative pomonoid structure
that interacts with the existing pomonoid via a kind of interchange
law (the \cref{rule:Sequence} rule in the proof system). We follow
\cite{hopf-book} generalising these to the case of monoids with
differing units. Their terminology is of a category having duoidal
structure. We find it useful to describe one pomonoid as being
\emph{duoidal over} another to emphasise the non-symmetric nature of
the relationship, and by analogy with one monoid distributing over
another.

@@

\subsection{MAV-algebras}

@@

\begin{proposition}\label[prop]{prop:mav-algebra-consequences}
  Let $(A, \leq, \otimes, \lhd, I, \lnot)$ be a MAV-algebra.
  \begin{enumerate}
    \item There is another commutative pomonoid structure $(\parr, I)$ on
          $(A, \leq)$, defined as
          $x \parr y = \lnot(\lnot x \otimes \lnot y)$.
    \item $(\otimes, I)$ and $(\parr, I)$ are linearly distributive
          \cite{cockett}:
          $x \otimes (y \parr z) \leq (x \otimes y) \parr z$.
    \item $(A, \leq)$ has binary joins, given by
          $x \oplus y = \lnot (\lnot x \with \lnot y)$.
    \item $\oplus$ distributes over $\otimes$:
          $x \otimes (y \oplus z) = (x \otimes y) \oplus (x \otimes z)$.
    \item $\with$ distributes over $\parr$:
          $(x \parr z) \with (y \parr z) = (x \with y) \parr z$.
    \item $\lhd$ is duoidal over $\parr$:
          $(w \parr x)\lhd (y \parr z) \leq (w\lhd y) \parr (x\lhd z)$.
    \item $\lhd$ is duoidal over $\with$:
          $(w \with x)\lhd (y \with z) \leq (w \lhd y) \with (x\lhd z)$.
    \item $\oplus$ is duoidal over $\lhd$:
          $(w \lhd x) \oplus (y \lhd z) \leq (x \oplus y) \lhd (x \oplus z)$.
  \end{enumerate}
\end{proposition}

@@

To prove completeness of the normal proofs of MAV, we will construct a
particular MAV-algebra from the formulas and normal proofs. Since
normal proofs do not \emph{a priori} have all the necessary structure
for an MAV-algebra, in the following sections we develop a procedure
to construct MAV-algebra from the lighter requirements of an
MAV-frame. In \Cref{sec:mav-cut-elimination} we will show that
the MAV-algebra constructed from the normal proof MAV-frame allows us
to prove that all proofs in MAV can be normalised to normal proofs.

\begin{definition}\label[defn]{defn:mav-frame}
  An \emph{MAV-frame} is a structure $(F, \leq, \parr, \lhd, I, +)$
  where $(F, \leq)$ is a partial order, $(F, \leq, \parr, I)$ is a
  commutative pomonoid, $(F, \leq, \lhd, I)$ is a pomonoid, $+$ is a
  binary monotone function on $(F, \leq)$, and these data satisfy the
  following inequalities:
  \begin{enumerate}
    \item $(w \lhd x) \parr (y \lhd z) \leq (w \parr y) \lhd (x \parr z)$
    \item $(x + y) \parr z \leq (x \parr z) + (y \parr z)$
    \item $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$
    \item $I + I \leq I$
  \end{enumerate}
\end{definition}

@@

\begin{remark}
  This is the Day monoidal product \cite{day} restricted to the case
  of partial orders.
\end{remark}

@@

In the phase semantics, formulas are interpreted as elements that are
fixed points of a closure operator defined by a double negation with
respect to the monoid on the original frame. This closure operator
generates a partial order of ``facts'' whose meets and joins exactly
correspond to the syntactic ones when the original monoid is derived
from the proofs. As we mentioned in the introduction, the presence of
a self-dual operator in BV means that we cannot use double negation
closure, and we have to proceed more deliberately to preserve
join-like structure in an MAV-frame when building MAV-algebras. We do
this by defining $+$-closed lower sets as those that are closed under
finite $+$-combinations of their elements. This leads to a closure
operator on lower sets that allows us to immediately deduce that
$+$-closed lower sets form a lattice. We also preserve the
Day-pomonoids from lower sets, but in two different ways, depending on
how the original pomonoid interacts with $+$. In \Cref{prop:closed-monoid-distrib} we handle pomonoids that distribute
over $+$, and in \Cref{prop:closed-monoid-duoidal} we
handle pomonoids that are entopic under $+$. We need these
constructions to lift the $\parr$ and $\lhd$ pomonoids from
MAV-frames to $+$-closed lower sets. Finally in this section, we show
that duoidal structure on lower sets from
\Cref{prop:lower-set-duoidal} is preserved in $+$-closed lower sets.

@@

\begin{remark}
  By this proposition, $U \circ \alpha$ is a closure operator on
  $\LowerSet{A}$ \cite{davey-priestley}, and the closed elements are
  those of $\ClosedLowerSet{A}$. The next proposition is standard for
  showing that meets and joins exist on the closed elements for some
  closure operator.
\end{remark}

@@

\begin{remark}
  \Cref{prop:closed-eta-preserve-joins} is the reason for
  requiring $+$-closure. This property will allow us to prove the
  crucial embedding property for all formulas in \Cref{sec:mav-cut-elimination}.
\end{remark}

@@

To construct suitable MAV-algebras, we use the partial order version
of the Chu construction \cite{barr}. The Chu construction builds
$*$-autonomous categories from symmetric monoidal closed categories
with pullbacks. In the partial order case, the requirement for
pullbacks simplifies to binary meets. For this section, we let
$(A, \leq, \bullet, i, \rightblackspoon)$ be a partial order with a
residuated pomonoid structure and all binary meets.

@@

\begin{theorem}\label[thm]{thm:mavframe-soundness}
  MAV is sound for the MAV-frame semantics: if $P \longrightarrow^* I$
  then for all MAV-frames $F$, $\sem{I} \sqsubseteq \sem{P}$ in
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$.
\end{theorem}

>>>>

To our knowledge, the semantics of MAV have not been previously
studied in the literature. The rules are an extension of MALL, so an
\emph{MAV-algebra} will partly be a $*$-autonomous partial order
(\Cref{defn:star-autonomous}) with meets and joins. The additional
structure for seq satisfies the conditions of a duoidal category
\cite[Definition 6.1]{Aguiar_2010} (\Cref{defn:duoidal}). We show that
MAV is sound for MAV-algebras in \Cref{thm:soundness}.

@@

BV and MAV extend linear logic by adding a non-commutative pomonoid
structure that interacts with the existing pomonoid via a kind of
interchange law (the \cref{rule:Sequence} rule in the proof
system). We follow \cite[Definition 6.1]{Aguiar_2010} generalising
these to the case of monoids with differing units. Their terminology
is of a category having duoidal structure. We find it useful to
describe one pomonoid as being \emph{duoidal over} another to
emphasise the non-symmetric nature of the relationship, and by analogy
with one monoid distributing over another.

@@

\subsection{MAV-algebras}
\label{sec:mav-algebras}

@@

\begin{proposition}\label[prop]{prop:mav-algebra-consequences}
  Let $(A, \leq, \otimes, \lhd, I, \lnot)$ be a MAV-algebra.
  \begin{enumerate}
    \item There is another commutative pomonoid structure $(\parr, I)$ on
          $(A, \leq)$, defined as
          $x \parr y = \lnot(\lnot x \otimes \lnot y)$.
    \item $(\otimes, I)$ and $(\parr, I)$ are linearly distributive
          \cite{Cockett_1999}:
          $x \otimes (y \parr z) \leq (x \otimes y) \parr z$.
    \item $(A, \leq)$ has binary joins, given by
          $x \oplus y = \lnot (\lnot x \with \lnot y)$.
    \item $\oplus$ distributes over $\otimes$:
          $x \otimes (y \oplus z) = (x \otimes y) \oplus (x \otimes z)$.
    \item $\with$ distributes over $\parr$:
          $(x \parr z) \with (y \parr z) = (x \with y) \parr z$.
    \item $\lhd$ is duoidal over $\parr$:
          $(w \parr x)\lhd (y \parr z) \leq (w\lhd y) \parr (x\lhd z)$.
    \item $\lhd$ is duoidal over $\with$:
          $(w \with x)\lhd (y \with z) \leq (w \lhd y) \with (x\lhd z)$.
    \item $\oplus$ is duoidal over $\lhd$:
          $(w \lhd x) \oplus (y \lhd z) \leq (x \oplus y) \lhd (x \oplus z)$.
  \end{enumerate}
\end{proposition}

@@

To prove completeness of the normal proofs of MAV, we will construct a
particular MAV-algebra from the structures and normal proofs. Since
normal proofs do not \emph{a priori} have all the necessary structure
for an MAV-algebra, in the following sections we develop a procedure
to construct MAV-algebra from the lighter requirements of an
MAV-frame. In \Cref{sec:mav-cut-elimination} we will show that
the MAV-algebra constructed from the normal proof MAV-frame allows us
to prove that all proofs in MAV can be normalised to normal proofs.

\begin{definition}\label[defn]{defn:mav-frame}
  An \emph{MAV-frame} is a structure $(F, \leq, \parr, \lhd, i, +)$
  where $(F, \leq)$ is a partial order, $(F, \leq, \parr, i)$ is a
  commutative pomonoid, $(F, \leq, \lhd, i)$ is a pomonoid, $+$ is a
  binary monotone function on $(F, \leq)$, and these data satisfy the
  following inequalities:
  \begin{enumerate}
    \item $(w \lhd x) \parr (y \lhd z) \leq (w \parr y) \lhd (x \parr z)$
    \item $(x + y) \parr z \leq (x \parr z) + (y \parr z)$
    \item $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$
    \item $i + i \leq i$
  \end{enumerate}
\end{definition}

@@

\begin{remark}
  \Cref{prop:day-construction} is the Day monoidal product on functor
  categories \cite{Day_1970} restricted to the case of partial orders
  and lower sets.
\end{remark}

@@

In the phase semantics, structures are interpreted as elements that are
fixed points of a closure operator defined by a double negation with
respect to the monoid on the original frame. This closure operator
generates a partial order of ``facts'' whose meets and joins exactly
correspond to the syntactic ones when the original monoid is derived
from the proofs. As we mentioned in the introduction, the presence of
a self-dual operator in BV means that we cannot use double negation
closure, and we have to proceed more deliberately to preserve
join-like structure in an MAV-frame when building MAV-algebras. We do
this by defining $+$-closed lower sets as those that are closed under
finite $+$-combinations of their elements. This leads to a closure
operator on lower sets that allows us to immediately deduce that
$+$-closed lower sets form a lattice. We also preserve the
Day-pomonoids from lower sets, but in two different ways, depending on
how the original pomonoid interacts with $+$. In \Cref{prop:closed-monoid-distrib} we handle pomonoids that distribute
over $+$, and in \Cref{prop:closed-monoid-duoidal} we
handle pomonoids that are entopic under $+$. We need these
constructions to lift the $\parr$ and $\lhd$ pomonoids from
MAV-frames to $+$-closed lower sets. Finally in this section, we show
that duoidal structure on lower sets from
\Cref{prop:lower-set-duoidal} is preserved in $+$-closed lower sets.

@@

\begin{remark}
  By this proposition, $U \circ \alpha$ is a closure operator on
  $\LowerSet{A}$ \cite{Davey_2002}, and the closed elements are
  those of $\ClosedLowerSet{A}$. The next proposition is standard for
  showing that meets and joins exist on the closed elements for some
  closure operator.
\end{remark}

@@

\begin{remark}
  \Cref{prop:closed-eta-preserve-joins} is the reason for
  requiring $+$-closure. This property will allow us to prove the
  crucial embedding property for all structures in \Cref{sec:mav-cut-elimination}.
\end{remark}

@@

To construct suitable MAV-algebras, we use the partial order version
of the Chu construction \cite[Appendix by Po-Hsiang
Chu]{Barr_1979}. The Chu construction builds $*$-autonomous categories
from symmetric monoidal closed categories with pullbacks. In the
partial order case, the requirement for pullbacks simplifies to binary
meets. For this section, we let
$(A, \leq, \bullet, i, \rightblackspoon)$ be a partial order with a
residuated pomonoid structure and all binary meets.

@@

\begin{theorem}\label[thm]{thm:mavframe-soundness}
  MAV is sound for the MAV-frame semantics: if $P \vInferFrom* \vUnit$
  then for all MAV-frames $F$, $\sem{I} \sqsubseteq \sem{P}$ in
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$.
\end{theorem}
